Software reliability: Experiences\\ in European scientific research\\ projects and new trends

### 2017-11-08 paper.tex
shortened to 3 pages

The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted to develop software to provide Grid computing functionalities and related management tools for widely distributed scientific communities, such as the Large Hadron Collider (LHC) experiments, Earth observations and Bioinformatics. In this project, a large effort was dedicated in the definition and implementation of academic software engineering practices due to the unusual context in which DataGrid aims, starting from the large geographical distribution of the teams involved in the development tasks, evolving requirements coming from the different application areas and the design of a new technology: the Grid \cite{datagrid}. DataGrid developed its own software distribution, {\sl EDG} (European DataGrid), strongly based on Globus Middleware Services \cite{globus}. An Architecture Task Force coordinated the overall design and technical consistency of the developments, complemented by the definition of various testing phases: preparation, execution and documentation.

The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science, available 24 hours--a--day. One of the main achievements of these projects was the middleware -- a crucial component of any Grid infrastructure as it provides the {\sl glue} that links the hardware resources within such infrastructure. The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts to converge with LHC Computing Grid ({\sl LCG-2}), Virtual Data Toolkit ({\sl VDT}) and {\sl Condor} \cite{condor} software distributions. The overall goal of the middleware engineering group was to provide and maintain selected middleware services of the {\sl gLite} distribution to satisfy the needs of the scientific communities. During EGEE--I, a software process was designed and put into action for the early releases of the {\sl gLite} middleware. This release process favoured a development environment of rapidly changing interfaces that were necessary during re--engineering. Starting in EGEE--II, there was a need to move to a more dynamic release process, and the first steps were taken towards an automated environment. At this stage  no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service. The {\sl gLite} middleware developed within EGEE will evolve further as part of a new project: the European Middleware Initiative (EMI).

@@

The EMI (May 2010 -- Apr 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe -- {\sl ARC}, {\sl dCache}, {\sl gLite} and {\sl UNICORE}. The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics that need to be present in EMI software products and processes to be able to meet EMI quality requirements. For each software characteristic, a set of associated metrics and Key Performance Indicators (KPIs) were identified and defined in detail in the EMI Metrics Specification \cite{emi-metrics}. EMI was one of the first projects in using a Continuous Integration (CI) system for building, packaging and testing the various components. It leveraged the {\sl ETICS} tool and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{images/unittest.png}
\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
\label{fig:fig_unittest}
\end{figure*}

@@

The requirements are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly. The source code is publicly available in GitHub repositories under an organization called {\sl indigo-dc} \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption. In this regard, the source code is compliant with community de--facto style standards, selected from the variety of programming languages being used.

@@

Tests are triggered in each build to measure the unit and functional testing coverage, with a recommended value of 70\% of coverage in the former case. The documentation of the products is treated as source code, using a markup language, automatically rendered and uploaded in online repositories \cite{indigo-gitbook}. Changes in both documentation and source code are human--reviewed as the last step before merging them into the production repository. Last but not least, as already seen in the DevOps section, in order to facilitate the usage of the INDIGO--DataCloud services, they can be deployed automatically using either Ansible \cite{indigo-ansible} or Puppet \cite{indigo-puppet} open--source tools.

The evaluation of the software quality is performed by measuring the values of the metrics and Key Performance Indicators (KPIs) defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard. These metrics cover the development, release and maintenance phases of the software lifecycle. Development and release metrics are obtained automatically from several sources, such as GitHub and Jenkins CI service, and graphically displayed as GitHub pages using GrimoireLab framework \cite{grimoirelab}. Maintenance and user support metrics are collected from the different sources of data like GGUS \cite{ggus} and Github Issues.

@@

\bibitem{cordis:datagrid}
\emph{DataGrid project}, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/53665\_en.html

\bibitem{datagrid}
[3] L. Momtahan, A. Martin, \emph{e-Science Experiences: Software Engineering Practice and the EU DataGrid}, in Proc. Asia-Pacific Software Engineering Conference, Gold Coast, Queensland, Australia, 4-6 Dec. 2002, pp. 269-275, IEEE Press

@@

\bibitem{cordis:egee}
\emph{Enabling Grids for E-sciencE (EGEE)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/80149\_en.html

\bibitem{cordis:egee2}
\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/99189\_en.html

\bibitem{cordis:egee3}
\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/87264\_en.html

\bibitem{glite}
E. Laure et al, \emph{Programming the Grid with gLite}, in Jin, H., Reed, D.A., Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1), pp. 33–45, 2006, Scientific Publishers OWN

\bibitem{condor}
D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the grid}, in Grid Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70, 2002, Eds. John Wiley \& Sons Inc.

\bibitem{cordis:etics}
\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/80138\_en.html

\bibitem{cordis:etics2}
\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/86604\_en.html

\bibitem{etics}
A. Di Meglio, M.-E.Bégin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol. 119, N. 4, 2008, IOP Publishing Ltd

\bibitem{cordis:emi}
\emph{European Middleware Initiative (EMI)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/95311\_en.html

\bibitem{iso9126}
\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International Organization for Standarization. Available: https://www.iso.org/standard/22749.html

\bibitem{emi-metrics}
\emph{EMI Metrics Specification}. Available: https://goo.gl/LdS6fL

\bibitem{cordis:egi-inspire}
\emph{European Grid Initiative: Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/95923\_en.html

\bibitem{egi-qc}
\emph{EGI Quality Criteriai}. Available: http://egi-qc.github.io/

@@

\bibitem{cordis:egi-engage}
\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/194937\_en.html

\bibitem{cordis:indigo}
\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/194882\_en.html

\bibitem{indigo-dockerhub}
\emph{INDIGO-DataCloud DockerHub repository}. Available: https://hub.docker.com/u/indigodatacloud

\bibitem{indigo-d31}
\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1. Available: https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31

\bibitem{indigo-jenkins}
\emph{INDIGO-DataCloud Jenkins CI service}. Available: https://jenkins.indigo-datacloud.eu:8080/

\bibitem{indigo-github}
\emph{INDIGO-DataCloud GitHub Source Code repository}. Available: https://www.github.com/indigo-dc

\bibitem{indigo-gitbook}
\emph{INDIGO-DataCloud GitBook Documentation repository}. Available: https://www.gitbook.com/@indigo-dc

\bibitem{indigo-puppet}
\emph{INDIGO-DataCloud PuppetForge repository}. Available: https://forge.puppet.com/indigodc

\bibitem{indigo-ansible}
\emph{INDIGO-DataCloud Ansible Galaxy repository}. Available: https://galaxy.ansible.com/indigo-dc/

\bibitem{grimoirelab}
\emph{GrimoireLab}. Available: http://grimoirelab.github.io/

\bibitem{ggus}
\emph{Global Grid User Support (GGUS)}. Available: https://www.ggus.eu/

>>>>

The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted to develop software to provide Grid computing functionalities and related management tools for widely distributed scientific communities, such as the Large Hadron Collider (LHC) experiments, Earth observations and Bioinformatics. In this project, a large effort was dedicated in the definition and implementation of academic software engineering practices due to the unusual context in which DataGrid aims, starting from the large geographical distribution of the teams involved in the development tasks, evolving requirements coming from the different application areas and the design of a new technology: the Grid \cite{datagrid}. DataGrid developed its own software distribution, {\sl EDG} (European DataGrid), strongly based on Globus Middleware Services \cite{globus}.
%An Architecture Task Force coordinated the overall design and technical consistency of the developments, complemented by the definition of various testing phases: preparation, execution and documentation.

The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science. The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering effortsi.
%to converge with LHC Computing Grid ({\sl LCG-2}), Virtual Data Toolkit ({\sl VDT}) and {\sl Condor} \cite{condor} software distributions.
Starting in EGEE--II, there was a need to move to a more dynamic release process, and the first steps were taken towards an automated environment. At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service. The {\sl gLite} middleware developed within EGEE will evolve further as part of a new project: the European Middleware Initiative (EMI).

@@

The EMI (May 2010 -- Apr 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe -- {\sl ARC}, {\sl dCache}, {\sl gLite} and {\sl UNICORE}. The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics, each defining the associated metrics and Key Performance Indicators (KPIs). EMI was one of the first projects in using a Continuous Integration (CI) system for building, packaging and testing the various components. It leveraged the {\sl ETICS} tool and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.

@@

%\begin{figure*}
%\centering
%\includegraphics[width=\textwidth]{images/unittest.png}
%\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
%\label{fig:fig_unittest}
%\end{figure*}

@@

The requirements are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly. The source code is publicly available as GitHub repositories \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption. In this regard, the source code is compliant with community de--facto style standards, selected from the variety of programming languages being used. Tests are triggered for each change to measure the unit and functional testing coverage, with a recommended value of 70\% of coverage in the former case.

@@

The evaluation of the software quality is performed by measuring the values of the metrics and Key Performance Indicators (KPIs) defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard. These metrics cover the development, release and maintenance phases of the software lifecycle.

@@

\bibitem{cordis:datagrid}
\emph{DataGrid project}, CORDIS Europe, http://cordis.europa.eu/project/rcn/53665\_en.html

\bibitem{datagrid}
[3] L. Momtahan et al, \emph{e-Science Experiences: Software Engineering Practice and the EU DataGrid}, in Proc. Asia-Pacific Software Engineering Conference, 2002, pp. 269-275, IEEE Press

@@

\bibitem{cordis:egee}
\emph{Enabling Grids for E-sciencE (EGEE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html

\bibitem{cordis:egee2}
\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/99189\_en.html

\bibitem{cordis:egee3}
\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/87264\_en.html

\bibitem{glite}
E. Laure et al, \emph{Programming the Grid with gLite}, in Computational Methods in Science and Technology, vol. 12(1), pp. 33–45, 2006, Scientific Publishers OWN

\bibitem{cordis:etics}
\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80138\_en.html

\bibitem{cordis:etics2}
\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/86604\_en.html

\bibitem{etics}
A. Di Meglio et al, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol. 119, N. 4, 2008, IOP Publishing Ltd

\bibitem{cordis:emi}
\emph{European Middleware Initiative (EMI)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95311\_en.html

\bibitem{iso9126}
\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International Organization for Standarization, https://www.iso.org/standard/22749.html

\bibitem{cordis:egi-inspire}
\emph{Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95923\_en.html

\bibitem{egi-qc}
\emph{EGI Quality Criteria}, http://egi-qc.github.io/

@@

\bibitem{cordis:egi-engage}
\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194937\_en.html

\bibitem{cordis:indigo}
\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194882\_en.html

\bibitem{indigo-dockerhub}
\emph{INDIGO-DataCloud DockerHub}, https://hub.docker.com/u/indigodatacloud

\bibitem{indigo-d31}
\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1, https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31

\bibitem{indigo-jenkins}
\emph{INDIGO-DataCloud Jenkins}, https://jenkins.indigo-datacloud.eu:8080/

\bibitem{indigo-github}
\emph{INDIGO-DataCloud GitHub}, https://www.github.com/indigo-dc
